---
title: "Project Bike"
author: "Name Surname"
date: "2022-09-09"
output:
  pdf_document: default
  html_document: default
---

### Load packages
````{r}
library(data.table)
library(tidyverse)
library(janitor)
library(skimr)
library(lubridate)
````
### Upload and merge columns at once
````{r}
temp <- list.files ("C:/Users/Desktop/...", full.names = TRUE, pattern = "\\.csv$")
bike_rides <- rbindlist(lapply(temp, fread), fill = TRUE)
````
### Remove empty
````{r}
bike_rides <- janitor::remove_empty(bike_rides,which = c("cols"))
bike_rides <- janitor::remove_empty(bike_rides,which = c("rows"))
options(warn=-1)
# check whether there are any changes
dim(bike_rides)
````
### Change columns data type
````{r}
bike_rides$started_at <- lubridate::ymd_hms(bike_rides$started_at)
bike_rides$ended_at <- lubridate::ymd_hms(bike_rides$ended_at)

bike_rides$start_station_id <- as.integer(bike_rides$start_station_id)
bike_rides$end_station_id <- as.integer(bike_rides$end_station_id)
````
### Conveert blank string to NA and remove all
````{r}
bike_rides <- mutate(bike_rides, across(
where(is.character),~ if_else(.x=="", NA_character_, .x)))
bike_rides <- na.omit(bike_rides)
````
### Check for discrepancies
````{r}
unique(bike_rides$rideable_type)
unique(bike_rides$member_casual)
````


### Calculate ride length. Converting into minutes
````{r}
bike_rides$ride_length <- difftime(bike_rides$ended_at, bike_rides$started_at)
bike_rides$ride_length <- bike_rides$ride_length/60
bike_rides$ride_length <- round(bike_rides$ride_length,2)
````

### Ride length data type to numeric
````{r}
bike_rides$ride_length <- as.numeric(as.character(bike_rides$ride_length))
````

### Check for duplicates

* n_occur <- data.frame(table(bike_rides$ride_id))
* n_occur[n_occur$Freq >1,]
* bike_rides[bike_rides$ride_id %in% n_occur$Var1[n_occur$Freq > 1],]

### Remove duplicates 
````{r}
bike_rides <- bike_rides[!duplicated(bike_rides$ride_id)]
````

### Remove ride length less then 0
````{r}
bike_rides <- filter(bike_rides, ride_length > 0)
````

### Avg. ride length by day, week, year. Casual vs Member
````{r}
bike_rides$date <- as.Date(bike_rides$started_at)
bike_rides$month <- format(as.Date(bike_rides$date), "%B")
bike_rides$day <- format(as.Date(bike_rides$date), "%d")
bike_rides$year <- format(as.Date(bike_rides$date), "%Y")
bike_rides$day_of_week <- weekdays(bike_rides$date)
bike_rides$start_hr <- lubridate::hour(bike_rides$started_at)
````

### Order weekdays and months
````{r}
bike_rides$day_of_week <- ordered(bike_rides$day_of_week, levels = c("Sunday", "Monday",
                                                                     
  "Tuesday", "Wednesday","Thursday", "Friday", "Saturday"))

bike_rides$month <- ordered(bike_rides$month, levels=c("January","February","March",
                                                       
"April","May","June","July","August","September","October","November","December"))
````
### Average ride length by weekdays and member type
````{r}
aggregate(bike_rides$ride_length ~ bike_rides$member_casual + bike_rides$day_of_week,
          FUN = mean)
````
### Summary
````{r}
summary(bike_rides$ride_length)
```
### Comparing casual and member users
````{r}
bike_rides %>% 
  group_by(member_casual) %>% 
  summarise(avg_length=mean(ride_length), min_length=min(ride_length), 
  max_length=max(ride_length), median_length=median(ride_length))
````

### Rides by weekday and  type of a rider
````{r}
bike_rides %>% 
  group_by(day_of_week, member_casual) %>% 
  summarise(number_of_rides = n(), avg_duration=mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)
````
### New df for visualization
````{r}
options(scipen = 999)
viz_1 <- group_by(bike_rides, member_casual, day_of_week) %>% summarise(num_of_rides=n())
````

### viz. Weekdays by number of rides
````{r}
ggplot(viz_1, aes(x=day_of_week, y=num_of_rides, fill=member_casual))+
  geom_bar(stat='identity', position = "dodge")+ 
  labs(title = "Number of Rides by Day", subtitle = "April 2020 - March 2021" ,x="Weekdays", y="Number of Rides")+ 
  theme(axis.text.x = element_text(angle = 45),legend.title = element_blank())
````

### viz. Month
````{r}
viz_2 <- group_by(bike_rides, member_casual, month) %>% summarise(num_of_rides=n())

ggplot(viz_2,aes(x=month,y=num_of_rides, fill=member_casual))+
  geom_bar(stat = 'identity', position = 'dodge')+ 
  labs(title = 'Number of Rides by Month', subtitle = "April 2020 - March 2021",x= 'Month', y='Number of Rides')+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6),
  legend.title = element_blank())
````                                   

### Scatterplot. AVG ride length
````{r}
### Scatterplot. AVG ride length

scatterplot <- bike_rides %>%
  group_by(month, day_of_week, member_casual) %>%
  summarise(avg_ride_length = mean(ride_length)) 
scatterplot %>%
  ggplot(aes(x = day_of_week, y = avg_ride_length, color = member_casual))+
  geom_point()+ facet_wrap(~month)+
  theme(axis.text.x = element_text(angle = 45,vjust = 0.8),
  legend.title = element_blank())+
  labs(title = "Average Ride Length for Each Month", 
  subtitle = "April 2020 - March 2021",y="Ride Length", x="Weekdays")
```` 

### Popular start stations
````{r}
pop_start_stations <- bike_rides %>% 
  count(start_station_name, member_casual)
````

### Popular end stations
````{r}
pop_end_stations <- bike_rides %>% 
  count(end_station_name,member_casual)
````

### Top 10 popular stations. Casual
````{r}
pop_start_casual <- filter(pop_start_stations, 
                      member_casual=="casual")
pop_start_casual <- pop_start_casual %>%  
  arrange(desc(n)) %>% 
  slice_head(n=10)
print(pop_start_casual)
ggplot(pop_start_casual)+geom_col(mapping = aes(
  x=reorder(start_station_name,-n),y=n,fill=start_station_name))+ labs(
    title = "Top 10 Start Stations - Casual",
    subtitle = "April 2020 - March 2021", 
    x="Station Names", y= "Ride Count")+ 
  theme(legend.position = "none")+
  coord_flip()
````

### Top 10 popular stations. Member
````{r}
pop_start_member <- filter(pop_start_stations, 
                           member_casual=="member")
pop_start_member <- pop_start_member %>%  
  arrange(desc(n)) %>% 
  slice_head(n=10)
print(pop_start_member)
ggplot(pop_start_member)+geom_col(mapping = aes(
  x=reorder(start_station_name,-n),y=n, fill=start_station_name))+ labs(
    title = "Top 10 Start Stations - Member",
    subtitle = "April 2020 - March 2021", 
    x="Station Names", y= "Ride Count")+ 
  theme(legend.position = "none")+
  coord_flip()
````
